---
title: "day4-01-trajectory.Rmd"
author: "Doron Betel & Eddie Imada"
date: "2022-12-09"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Loading the libraries

```{r}
library(scater)
library(scran)
library(ggplot2)
```


# Load the Nestorowa et al dataset

```{r}
sce.nest <- readRDS("~/workspace/practical_genomics/day4/sce.nest.rds")
```

# Cluster based minimum spanning trees

The first step is to obtain the centroids for each cluster:
```{r}
by.cluster <- aggregateAcrossCells(sce.nest, ids=colLabels(sce.nest))
centroids <- reducedDim(by.cluster, "PCA")
```

Now we compute the minimum spanning tree (MST) with TSCAN. The MST is simply an 
undirected acyclic graph that passes through each centroid exactly once and is 
thus the most parsimonious structure that captures the transitions between 
clusters. 

There are multiple ways to create the MST with TSCAN. In this example we are
starting with the centroids coordinates in a reduce dimension (PCA). We can
use the function `createClusterMST()` to create the MST.

```{r}
library(TSCAN)
# We set cluster=NULL since we already aggregate the by cluster in the code above
mst <- createClusterMST(centroids, clusters=NULL)
mst
```

Next, use the function `reportEdges()` to obtain the coordinates for the start
and end of of every edge (line conecting two vertices) in the MST. We can
obtain these coordinates in spaces different from what was used to create the
MST. Here we want to visualize with TSNE and the MST was created in PCA space.

```{r}
line.data <- reportEdges(by.cluster, mst=mst, clusters=NULL, use.dimred="TSNE")
line.data
```

Now we can plot our data in the TSNE space and add the lines we just obtained.
For that we can use `geom_line()`from ggplot2.

```{r}
p <- plotTSNE(sce.nest, colour_by="label") + 
  geom_line(data=line.data, mapping=aes(x=dim1, y=dim2, group=edge))
print(p)
```

Now that we have the paths, we can compute the pseudotime for each cell.
For that first we use the function `mapCellToEdges()` to map each cell to the
nearest position on an edge.

```{r}
map.tscan <- mapCellsToEdges(sce.nest, mst=mst, use.dimred="PCA")
```

Next, we can compute the distance from the positions obtained above to the root
(the point in which the pseudotime is relative to). In this case we will choose
a random root for demonstration purposes. However, it is often more meaningful
to choose a root based on biology (i.e. a more undiferentiated/early state).

```{r}
tscan.pseudo <- orderCells(map.tscan, mst)
head(tscan.pseudo)
```

We can see we have 2 columns in our `tscan.pseudo` object. Each column reports
the pseudotime orderings from the root to each terminal node. Since we have a
branch in this example we have 2 terminal nodes (apart from the root).

Now we can compute the pseudotime with the `averagePseudotime()` function.

```{r}
common.pseudo <- averagePseudotime(tscan.pseudo) 
```

Let's visualize our data and color the cells by their pseudotime.

```{r}
p <- plotTSNE(sce.nest, colour_by=I(common.pseudo), 
              text_by="label", text_colour="red") +
  geom_line(data=line.data, mapping=aes(x=dim1, y=dim2, group=edge))
print(p)
```

# Principal curves

Another approach to computing pseudotime is using principal curves. Principal
curve is effectively a non-linear generalization of PCA, where the curve is
allowed to bend. 

To exercute that we can use the `slingshot` package and `slingshot()` function
fit a principal curve and compute the pseudotime similarly to the previous
approach (projecting into the curve and copmuting the distance to the root).

```{r}
library(slingshot)
sce.sling <- slingshot(sce.nest, reducedDim='PCA')
head(sce.sling$slingPseudotime_1)
```

We now have the pseudotimes for each cell. To visualize our data, we can
use the function `embedCurves()` to obtain the path the curve is taking in
any coordinate space we choose (TSNE in this example). Next, we can extract the
coordinates in this space with the `slingCurves()` function.

```{r}
embedded <- embedCurves(sce.sling, "TSNE")
embedded <- slingCurves(embedded)[[1]] # only 1 path.
```

Now we extract the coordinates in the space (in the `s` slot) and sort by the
order (in the `ord` slot).

```{r}
embedded <- data.frame(embedded$s[embedded$ord,])
```

Now we can plot the data and the curve with `geom_path()`. NOTE: We cannot use
`geom_line()` here since the principal curve is not linear.

```{r}
p <- plotTSNE(sce.sling, colour_by="slingPseudotime_1") +
  geom_path(data=embedded, aes(x=Dim.1, y=Dim.2), size=1.2)

print(p)
```

In this example we assumed that all cells are part of a single curve (no branch)
However, from our MST analysis we have data that suggests branching. We can
tweak `slingshot()` parameters to build a MST and use it to define the number
of "lineages". Each path from the root to a terminal node is considered one 
lineage. We can build a MST by passing the clusters ID to the `cluster`
parameter.

Moreover, by default `slingshot()` will compute one point of the curve per cell,
which can quickly become computationally demanding with a high number of cells.
We can reduce this load by requesting that slingshot approximate these points
by N cells with the parameter `approx_points`.

```{r}
sce.sling4 <- slingshot(sce.nest, cluster=colLabels(sce.nest), 
                        reducedDim='PCA', approx_points=100, omega=TRUE)
pseudo.paths <- slingPseudotime(sce.sling4)
head(pseudo.paths)
```

We can see we have 3 lineages in our example, however, we can see that some
cells are shared across lineages, but each has a different pseudotime depending
on the lineage. To obtain a single pseudotime, we can summarize the data.
In this case we will average the pseudotimes.

```{r}
shared.pseudo <- rowMeans(pseudo.paths, na.rm=TRUE)
```

Now let's plot the data with pseudotime. First we obtain the curve coordinates.

```{r}
embedded <- embedCurves(sce.sling4, "UMAP")
embedded <- slingCurves(embedded)
length(embedded)
```

In this case we have 3 curves, so we need to extract the coordinates for each
curve.

```{r}
line1 <- embedded[[1]]
line1 <- data.frame(line1$s[line1$ord,])
line2 <- embedded[[2]]
line2 <- data.frame(line2$s[line2$ord,])
line3 <- embedded[[3]]
line3 <- data.frame(line3$s[line3$ord,])
```

Now we can plot our data.

```{r}
gg <- plotUMAP(sce.sling4, colour_by=I(shared.pseudo)) + 
  geom_path(data=line1, aes(x=Dim.1, y=Dim.2), size=1.2) +
  geom_path(data=line2, aes(x=Dim.1, y=Dim.2), size=1.2) +
  geom_path(data=line3, aes(x=Dim.1, y=Dim.2), size=1.2)

print(gg)
```



