---
title: "day3-03-markers"
author: "Eddie Imada"
date: "2022-12-01"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Load the packages
```{r, include=FALSE}
library(scran)
library(scater)
```


## Load the dataset

We are using the PBMC4k dataset in this example.

```{r}
sce.pbmc <- readRDS("~/workspace/practical_genomics/sce.pbmc.rds")
```

## Performing comparisons across clusters
The function `scoreMarkers()` performs pairwise comparisons between clusters
and returns a list of dataframes containing several metrics for each cluster.
It requires as input a `singleCellExperiment` object and a vector with the
labels for each cell.

```{r}
marker.info <- scoreMarkers(sce.pbmc, colLabels(sce.pbmc))
marker.info
```

Let's check the metrics that are available.

```{r}
colnames(marker.info[[1]])
```

Let's see how this data frame looks like for the cluster 1.
```{r}
head(marker.info[[1]])
```

## Let's explore the clusters

First lets select the cluster 1.
```{r}
chosen <- marker.info[["1"]]
```
Now let's sort it by the metric of choice (mean AUC) and check the top genes.
```{r}
ordered <- chosen[order(chosen$mean.AUC, decreasing=TRUE),]
head(ordered[,1:4]) # showing basic stats only, for brevity.
```

## Looking at expression profiles of the top markers

We can use the `plotExpression()` function and pass the names of the top markers
to check their expression profiles across the clusters.

```{r}
plotExpression(sce.pbmc, features=head(rownames(ordered)), 
               x="label", colour_by="label")
```

## Exploring other ways to summarize the metrics

Each metric and summarizing approach has pros and cons that should be considered
during our analysis. There isn't the BEST metric, since it dependend on our
goals. 

Let's see what, if anything, changes when we sort by the `min.AUC` instead of `mean.AUC`.

```{r}
ordered <- chosen[order(chosen$min.AUC, decreasing=TRUE),]
head(ordered[,1:4]) # showing basic stats only, for brevity.
plotExpression(sce.pbmc, features=head(rownames(ordered)), 
               x="label", colour_by="label")
```
It seems most of the top genes are the same. However, the gene SPON2 is new in 
our top 6 list. Why?

## Exploring other metrics

Let's see which markers do we get when sorting by the logFC metric.

```{r}
ordered <- chosen[order(chosen$mean.logFC.detected, decreasing=TRUE),]
head(ordered[,1:4]) # showing basic stats only, for brevity.
```

That's a completely different list! Let's check how their expression profiles
looks like.

```{r}
plotExpression(sce.pbmc, features=head(rownames(ordered)), 
               x="label", colour_by="label")
```

The difference is quite striking! Let's practice more!

